Limitations on the Two-Infall Scenario in the Era of Large Stellar Age Catalogs

### 2025-04-02 src/tex/ms.tex

% user-defined commands
% \newcommand{\yes}{\textcolor{green}{\checkmark}\xspace}
% \newcommand{\meh}{\textcolor{black}{$\sim$}\xspace}
% \newcommand{\no}{\textcolor{red}{$\times$}\xspace}
\newcommand{\osuaffil}{Department of Astronomy, The Ohio State University, 140 W. 18th Ave, Columbus OH 43210, USA}
\newcommand{\ccappaffil}{Center for Cosmology and AstroParticle Physics, The Ohio State University, 191 W. Woodruff Ave., Columbus OH 43210, USA}
\newcommand{\aFe}{[$\alpha$/Fe]\xspace}
\newcommand{\mathXH}{{\rm [X/H]}}
\newcommand{\mathOH}{{\rm [O/H]}}
\newcommand{\mathFeH}{{\rm [Fe/H]}}
\newcommand{\mathOFe}{{\rm [O/Fe]}}
\newcommand{\mathaFe}{{\rm [$\alpha$/Fe]}}
\newcommand{\yZ}[1]{$y/Z_\odot=#1$}
\newcommand{\vice}{{\tt VICE}\xspace}
% \newcommand{\hydro}{{\tt h277}\xspace}
\newcommand{\todo}[1]{{\color{red}#1}}

@@

\begin{table*}
    \centering
    \caption{Sample selection parameters and median uncertainties from APOGEE DR17 (see Section \ref{sec:observational-sample}).}
    \label{tab:sample}
    \begin{tabular}{lll}
        \hline\hline
        \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Range or Value} & \multicolumn{1}{c}{Notes} \\
        \hline
        $\log g$            & $1.0 < \log g < 3.8$          & Select giants only \\
        $T_{\rm eff}$       & $3500 < T_{\rm eff} < 5500$ K & Reliable temperature range \\
        $S/N$               & $S/N > 80$                    & Required for accurate stellar parameters \\
        ASPCAPFLAG Bits     & $\notin$ 23                   & Remove stars flagged as bad \\
        EXTRATARG Bits      & $\notin$ 0, 1, 2, 3, or 4     & Select main red star sample only \\
        Latent age          & $\sigma_\tau/\tau < 40\%$     & Age uncertainty from \citet{leung_variational_2023} \\
        $R_{\rm gal}$     & $3 < R_{\rm gal} < 15\,{\rm kpc}$    & Eliminate bulge \& extreme outer-disk stars \\
        $|z|$               & $|z| < 2\,{\rm kpc}$                 & Eliminate halo stars \\
        \hline
    \end{tabular}
\end{table*}

@@

\begin{table*}
    \centering
    \caption{Median and dispersion in APOGEE parameter uncertainties.}
    \label{tab:uncertainties}
    \begin{tabular}{lcc}
        \hline\hline
        \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Median Uncertainty} & \multicolumn{1}{c}{Uncertainty Dispersion ($95\%-5\%$)} \\
        \hline
        [Fe/H]          & $0.0089$   & $0.0060$ \\
        $\mathOFe$  & $0.019$    & $0.031$ \\
        $\log_{10}(\tau_{\rm latent}/{\rm Gyr})$    & 0.10   & $0.16$ \\
        $\tau_{\rm [C/N]}/{\rm Gyr}$     & 1.0   & \todo{N/A} \\
        \hline
    \end{tabular}
\end{table*}

@@

We supplement the APOGEE DR17 abundance data with two different age catalogs. The first is the ``latent'' age estimates from \citet{leung_variational_2023}, who train a variational encoder-decoder network on asteroseismic data for APOGEE red giants. This catalog has two main advantages for our purposes: their method is designed to reduce contamination from abundance information (in particular [$\alpha$/Fe]), and the recovered ages do not plateau at $\sim10\,{\rm Gyr}$ as they do in some other neural network-derived age catalogs \citep[e.g.,][]{mackereth_dynamical_2019}. Following the recommendations of \citet{leung_variational_2023}, we use an upper limit on the relative age uncertainty of 40\%. This produces a sample of \num{57607} stars with latent age estimates, of which \num{14871} are in the Solar neighborhood ($7\leq R_{\rm gal}<9\,{\rm kpc}$, $0\leq|z|<0.5\,{\rm kpc}$). The median uncertainty in $\log({\rm age}/{\rm Gyr})$ is 0.10 (see Table \ref{tab:uncertainties}).

@@

\begin{deluxetable*}{Cccl}
    \tablecaption{A summary of variables and their fiducial values for our chemical evolution models (see discussion in Section \ref{sec:methods}).\label{tab:multizone-parameters}}
    \tablehead{
        \colhead{Quantity} & \colhead{Fiducial Value(s)} & \colhead{Section} & \colhead{Description}
    }
    \startdata
        % Yields
        % y/Z_\odot       & 1         & \ref{sec:yields}  & Nucleosynthetic yield scaling (see Table \ref{tab:yields}) \\
        % N_{\rm Ia}/M_\star  & $1.55\times10^{-3}$       & \ref{sec:yields}  & Integrated Type Ia supernova rate (see Table \ref{tab:yields}) \\
        f_{\rm Ia}(t)   & Equation \ref{eq:plateau-dtd}   & \ref{sec:yields}  & Delay-time distribution of Type Ia supernovae \\
        % t_D             & 40 Myr    & \ref{sec:yields}  & Minimum SN Ia delay time \\
        % Star formation histories
        % Outflows
        % \eta_\odot      & 0.2       & \ref{sec:outflows}    & Outflow mass-loading factor in the Solar annulus ($\eta\equiv \dot\Sigma_{\rm out}/\dot\Sigma_\star$) \\
        R_\eta          & 5.0 kpc   & \ref{sec:outflows}    & Exponential outflow scale radius \\
        f_{\Sigma,0}    & 0.27      & \ref{sec:sfh}     & Central thick/thin disk surface density ratio \\
        \mathXH_{\rm CGM}   & Pristine  & \ref{sec:sfh}     & Metallicity of infalling gas \\
        \tau_1          & 1 Gyr     & \ref{sec:parameter-selection}     & Timescale of the first infall epoch \\
        \tau_2(R_\odot)  & 15 Gyr    & \ref{sec:parameter-selection}     & Timescale of the second infall epoch at the Solar annulus \\
        R_{\tau_2}   & 7 kpc     & \ref{sec:parameter-selection}      & Exponential scale radius of the second infall timescale \\
        t_{\rm max}      & 4.2 Gyr  & \ref{sec:parameter-selection}     & Time of maximum gas infall (onset of second infall) \\
        \sigma_{\rm RM8}    & 2.68 kpc  & \ref{sec:migration}   & Radial migration strength
    \enddata
\end{deluxetable*}
\vspace{-24pt}

@@

The population-averaged nucleosynthetic yields of CCSNe are still fairly uncertain. This problem is exacerbated by the complexity of the CCSN explosion landscape \citep{sukhbold_core-collapse_2016}. Recently, \citet{weinberg_scale_2024} used a measurement of the mean Fe yield of CC SNe by \citet{rodriguez_iron_2023} and the plateau in stellar \aFe abundances at low metallicity to infer population-averaged yields of $y/Z_\odot\approx1$ --- in other words, the scale of SN yields is similar to the Solar abundance. However, \citet{johnson_milky_2024} found that GCE models with yields at this scale approach present-day abundances too slowly. We therefore investigate yield sets at multiple scales, which are presented in Table \ref{tab:yields}.

The SN Ia yield of Fe, $y_{\rm Fe}^{\rm Ia}$, is calculated so that the combined Fe yield $y_{\rm Fe}^{\rm Ia+CC}=y/Z_\odot$. We further increase $y_{\rm Fe}^{\rm Ia}$ by a factor of 30\% and 10\% for the \yZ{1} and \yZ{2} yield sets, respectively, to ensure that our models reach $\mathOFe\approx0.0$ by the end time. The fifth row of Table \ref{tab:yields} reports the integrated SN Ia rate $N_{\rm Ia}/M_\star$ from each yield set, assuming a mean Fe yield per SN Ia of $\overline m_{\rm Fe}^{\rm Ia}=0.7$ M$_\odot$ \citep{mazzali_common_2007,howell_effect_2009}. The rate for the \yZ{1} yield set is slightly higher than the volumetric rate of $N_{\rm Ia}/M_\star=(1.3\pm0.1)\times10^{-3}\,{\rm M}_\odot^{-1}$ reported by \citet{maoz_star_2017}, but is consistent with their measurement of $N_{\rm Ia}/M_\star=(1.6\pm0.3)\times10^{-3}\,{\rm M}_\odot^{-1}$ for field galaxies. The rate for the \yZ{2} yield set is consistent with the measurement of $N_{\rm Ia}/M_\star=(2.2\pm1.0)\times10^{-3}\,{\rm M}_\odot^{-1}$ by \citet{maoz_type-ia_2012}, while the rate for the \yZ{3} yield set is generally higher than literature values.

@@

The discrepancies between the predicted and observed abundance evolution discussed in Section \ref{sec:age-abundance} are also seen across the disk. Figure \ref{fig:mdf-evolution} shows the evolution of the MDF with age across five radial bins. We select the $y/Z_\odot=1$ and $y/Z_\odot=2$ models with pre-enriched infall due to the relative success of these models in the previous sections. We use the [C/N]-derived age estimates for the APOGEE sample due to the larger sample size in the most distant regions of the disk; we limit the comparison to ages $\leq10$ Gyr because of large systematic uncertainties for the oldest stars, as discussed in Section \ref{sec:observational-sample}. The two-infall model predicts a very broad MDF for ages $>10$ Gyr because of the rapid enrichment of the ISM during the first infall epoch.

>>>>

% user-defined commands
% \newcommand{\yes}{\textcolor{green}{\checkmark}\xspace}
% \newcommand{\meh}{\textcolor{black}{$\sim$}\xspace}
% \newcommand{\no}{\textcolor{red}{$\times$}\xspace}
\newcommand{\osuaffil}{Department of Astronomy, The Ohio State University, 140 W. 18th Ave, Columbus OH 43210, USA}
\newcommand{\ccappaffil}{Center for Cosmology and AstroParticle Physics, The Ohio State University, 191 W. Woodruff Ave., Columbus OH 43210, USA}
\newcommand{\aFe}{[$\alpha$/Fe]\xspace}
\newcommand{\mathXH}{{\rm [X/H]}}
\newcommand{\mathOH}{{\rm [O/H]}}
\newcommand{\mathFeH}{{\rm [Fe/H]}}
\newcommand{\mathOFe}{{\rm [O/Fe]}}
\newcommand{\mathaFe}{[\alpha/{\rm Fe}]}
\newcommand{\yZ}[1]{$y/Z_\odot=#1$}
\newcommand{\vice}{{\tt VICE}\xspace}
% \newcommand{\hydro}{{\tt h277}\xspace}
\newcommand{\todo}[1]{{\color{red}#1}}

@@

\begin{table*}
    \centering
    \caption{Sample selection parameters from APOGEE DR17 (see Section \ref{sec:observational-sample}).}
    \label{tab:sample}
    \begin{tabular}{lll}
        \hline\hline
        \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Range or Value} & \multicolumn{1}{c}{Notes} \\
        \hline
        $\log g$            & $1.0 < \log g < 3.8$          & Select giants only \\
        $T_{\rm eff}$       & $3500 < T_{\rm eff} < 5500$ K & Reliable temperature range \\
        $S/N$               & $S/N > 80$                    & Required for accurate stellar parameters \\
        ASPCAPFLAG Bits     & $\notin$ 23                   & Remove stars flagged as bad \\
        EXTRATARG Bits      & $\notin$ 0, 1, 2, 3, or 4     & Select main red star sample only \\
        NN age error        & $\sigma_\tau/\tau < 40\%$     & Age uncertainty from \citet{leung_variational_2023} \\
        $R_{\rm gal}$     & $3 < R_{\rm gal} < 15\,{\rm kpc}$    & Eliminate bulge \& extreme outer-disk stars \\
        $|z|$               & $|z| < 2\,{\rm kpc}$                 & Eliminate halo stars \\
        \hline
    \end{tabular}
\end{table*}

@@

\begin{table*}
    \centering
    \caption{Median and dispersion in APOGEE parameter uncertainties.}
    \label{tab:uncertainties}
    \begin{tabular}{lcc}
        \hline\hline
        \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Median Uncertainty} & \multicolumn{1}{c}{Uncertainty Dispersion ($95\%-5\%$)} \\
        \hline
        [Fe/H]          & $0.0089$   & $0.0060$ \\
        $\mathOFe$  & $0.019$    & $0.031$ \\
        $\log_{10}(\tau_{\rm NN}/{\rm Gyr})$    & 0.10   & $0.16$ \\
        $\tau_{\rm [C/N]}/{\rm Gyr}$     & 1.0   & \todo{N/A} \\
        \hline
    \end{tabular}
\end{table*}

@@

We supplement the APOGEE DR17 abundance data with two different age catalogs. The first is from \citet{leung_variational_2023}, who train a variational encoder-decoder network on asteroseismic data for APOGEE red giants. This catalog has two main advantages over other neural network (NN) age estimates: their method is designed to reduce contamination from abundance information (in particular [$\alpha$/Fe]), and the recovered ages do not plateau at $\sim10\,{\rm Gyr}$ as they do in some other neural network-derived age catalogs \citep[e.g.,][]{mackereth_dynamical_2019}. Following the recommendations of \citet{leung_variational_2023}, we use an upper limit on the relative age uncertainty of 40\%. This produces a sample of \num{57607} stars with NN age estimates, of which \num{14871} are in the Solar neighborhood ($7\leq R_{\rm gal}<9\,{\rm kpc}$, $0\leq|z|<0.5\,{\rm kpc}$). The median uncertainty in $\log({\rm age}/{\rm Gyr})$ is 0.10 (see Table \ref{tab:uncertainties}). We hereafter refer to the \citet{leung_variational_2023} ages as ``NN ages.''

@@

\begin{deluxetable*}{Ccccl}
    \tablecaption{A summary of variables and their fiducial values for our chemical evolution models (see discussion in Section \ref{sec:methods}).\label{tab:multizone-parameters}}
    \tablehead{
        \colhead{Quantity} & \colhead{Fiducial Value(s)} & \colhead{Alternatives} & \colhead{Section} & \colhead{Description}
    }
    \startdata
        f_{\rm Ia}(t)       & Equation \ref{eq:plateau-dtd} & Equation \ref{eq:powerlaw-dtd}    & \ref{sec:yields}  & Delay-time distribution of Type Ia supernovae \\
        R_\eta              & 5.0 kpc   & ---               & \ref{sec:outflows}            & Exponential outflow scale radius \\
        f_\Sigma(R_\odot)   & 0.12      & 0.25, 0.5         & \ref{sec:sfh}                 & Local thick/thin disk surface density ratio \\
        \mathXH_{\rm CGM}   & Pristine  & $-0.7$, $-0.5$    & \ref{sec:sfh}                 & Metallicity of infalling gas \\
        \tau_1              & 1 Gyr     & $0.1-3$ Gyr       & \ref{sec:parameter-selection} & Timescale of the first infall epoch \\
        \tau_2(R_\odot)     & 15 Gyr    & $3-30$ Gyr        & \ref{sec:parameter-selection} & Timescale of the second infall epoch at the Solar annulus \\
        R_{\tau_2}          & 7 kpc     & ---               & \ref{sec:parameter-selection}      & Exponential scale radius of the second infall timescale \\
        t_{\rm max}         & 4.2 Gyr   & $1-5$ Gyr         & \ref{sec:parameter-selection}     & Time of maximum gas infall (onset of second infall) \\
        \sigma_{\rm RM8}    & 2.68 kpc  & 3.6, 5.0 kpc      & \ref{sec:migration}   & Radial migration strength
    \enddata
\end{deluxetable*}
\vspace{-24pt}

@@

The population-averaged nucleosynthetic yields of CCSNe are still fairly uncertain. This problem is exacerbated by the complexity of the CCSN explosion landscape \citep{sukhbold_core-collapse_2016}. Recently, \citet{weinberg_scale_2024} used a measurement of the mean Fe yield of CC SNe by \citet{rodriguez_iron_2023} and the plateau in stellar \aFe abundances at low metallicity to infer population-averaged yields of $y/Z_\odot\approx1$\textemdash in other words, the scale of SN yields is similar to the Solar abundance. However, \citet{johnson_milky_2024} found that GCE models with yields at this scale approach present-day abundances too slowly to match the observed age--metallicity relation. Previous multi-zone models using {\tt VICE} \citep[e.g.,][]{johnson_stellar_2021,dubay_galactic_2024} adopted higher yields based on \citet{chieffi_explosive_2004} and \citet{limongi_nucleosynthesis_2006} ($y/Z_\odot\approx2.6$).

We investigate yield sets at multiple scales of the Solar abundance. The CCSN yield of O is directly set by the Solar scale: $y_{\rm O}^{\rm CC}=(y/Z_\odot)Z_{\rm O,\odot}$. For Fe, the CCSN yield is set by the \aFe ``plateau'' at low metallicity, $\mathaFe_{\rm CC}$, such that $y_{\rm Fe}^{\rm CC}=(y/Z_\odot) Z_{\rm Fe,\odot} 10^{-\mathaFe}$ \citep[for further discussion on the empirical yield scale and the CCSN plateau, see][]{weinberg_scale_2024}. We adopt the \citet{asplund_chemical_2009} Solar abundances: $Z_{\rm O,\odot}=5.72\times10^{-3}$ and $Z_{\rm Fe,\odot}=1.29\times10^{-3}$. Our yield sets are presented in Table \ref{tab:yields}.

The SN Ia yield of Fe, $y_{\rm Fe}^{\rm Ia}$, is calculated so that the combined Fe yield $y_{\rm Fe}^{\rm Ia+CC}=y/Z_\odot$. We further increase $y_{\rm Fe}^{\rm Ia}$ by a factor of 30\% and 10\% for the \yZ{1} and \yZ{2} yield sets, respectively, to ensure that our models reach $\mathOFe\approx0.0$ by $t=13.2\,{\rm Gyr}$. The fifth row of Table \ref{tab:yields} reports the integrated SN Ia rate $N_{\rm Ia}/M_\star$ from each yield set, assuming a mean Fe yield per SN Ia of $\overline m_{\rm Fe}^{\rm Ia}=0.7$ M$_\odot$ \citep{mazzali_common_2007,howell_effect_2009}. The rate for the \yZ{1} yield set is slightly higher than the volumetric rate of $N_{\rm Ia}/M_\star=(1.3\pm0.1)\times10^{-3}\,{\rm M}_\odot^{-1}$ reported by \citet{maoz_star_2017}, but is consistent with their measurement of $N_{\rm Ia}/M_\star=(1.6\pm0.3)\times10^{-3}\,{\rm M}_\odot^{-1}$ for field galaxies. The rate for the \yZ{2} yield set is consistent with the measurement of $N_{\rm Ia}/M_\star=(2.2\pm1.0)\times10^{-3}\,{\rm M}_\odot^{-1}$ by \citet{maoz_type-ia_2012}, while the rate for the \yZ{3} yield set is generally higher than literature values.

@@

The discrepancies between the predicted and observed abundance evolution discussed in Section \ref{sec:age-abundance} are also seen across the disk. Figure \ref{fig:mdf-evolution} shows the evolution of the MDF with age across five radial bins for the \yZ{1} and \yZ{2} models. We adopt pre-enriched infall ($\mathFeH_{\rm CGM}=-0.7$), stronger radial migration ($\sigma_{\rm RM8}=3.6$ kpc), and a higher disk mass ratio ($f_\Sigma(R_\odot)=0.25$) because these parameters were able to improve agreement with observations in previous sections. For the APOGEE sample, we use the [C/N]-derived age estimates due to the larger sample size in the most distant regions of the disk; we limit the comparison to ages $\leq10$ Gyr because of large systematic uncertainties for the oldest stars, as discussed in Section \ref{sec:observational-sample}.

The two-infall model predicts a very broad MDF for ages $>10$ Gyr because of the rapid enrichment of the ISM during the first infall epoch.
